---
title: "Credit Portfolio Risk Model"
author: "Alexander Zhuk"
date: "2026-07-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Part I: Portfolio Generation & Expected Loss

Building a synthetic corporate loan portfolio with S&P-style ratings,
assigning PD / LGD / EAD, and computing Expected Loss (EL = PD x LGD x EAD).

```{r cars}
set.seed(42) #for reproducibility
```

### 1. PD by rating

Approximated long-run 1-year corporate default rates (S&P-style rating scale).
PDs are anchored to published S&P historical default rates.

```{r card}
pd_table <- data.frame(
rating = c("AAA", "AA", "A", "BBB", "BB", "B", "CCC"),
pd = c(0.0001, 0.0002, 0.0006, 0.0018, 0.0075, 0.0380, 0.2500)
)
```

Portfolio mix: mostly investment grade, for a regional bank.

```{r}
rating_weights <- c(AAA=0.02, AA=0.08, A=0.20, BBB= 0.35, BB=0.20, B=0.12, CCC=0.03) # the portfolio mix
```

### 2. LGD by segment

Recovery depends on collateral and industry. LGD = 1 - recovery rate.
Ranges reflect typical assumptions (25%-90%).

```{r}
segment_table <- data.frame(
segment=c("CRE_secured","Manufacturing","Corporate_unsecured", "Software"),
lgd_min=c(0.25, 0.35, 0.50, 0.75),
lgd_max=c(0.40, 0.55, 0.70, 0.90),
weight=c(0.35, 0.25, 0.25, 0.15)
)
```

### 3. Generating the portfolio

```{r}
generate_portfolio <- function(n=500) {

rating <- sample(names(rating_weights), n, replace = TRUE, prob= rating_weights)
segment <- sample(segment_table$segment, n, replace = TRUE, prob = segment_table$weight)
#Committed amount: lognormal so a few large loans dominate(realistic)
committed <- round(rlnorm(n, meanlog = log(2e6), sdlog = 0.9), -3)


#Facility type: term loans are fully drawn: revolvers are partially drawn


facility <- sample(c("term", "revolver"), n, replace = TRUE, prob = c(0.6, 0.4))
drawn_pct <-ifelse(facility== "term", 1.0, runif(n, 0.2, 0.6))



#EAD: drawn amount + credit conversion factor (CCF) on the undrawn portion.
#revolvers get drawn down heavily near default -> CCF = 75%

ccf <- 0.75
ead <- committed * drawn_pct + committed * (1-drawn_pct)*ccf

#look up PD by rating
pd<- pd_table$pd[match(rating,pd_table$rating)]

#Draw LGD uniformly within segment's range
seg_idx <-match(segment, segment_table$segment)
lgd<- runif(n,segment_table$lgd_min[seg_idx], segment_table$lgd_max[seg_idx])

data.frame(
loan_id = sprintf("L%04d", 1:n),
rating, segment, facility,
committed, ead=round(ead),
pd,lgd = round(lgd, 3),
expected_loss=round(pd*lgd*ead)
)
}

portfolio <- generate_portfolio(500)
```

### 4. Portfolio summaries

```{r}
cat("===Portfolio Review===\n")
cat("Loans:", nrow(portfolio),
    "| Total EAD: $", format(sum(portfolio$ead), big.mark = ","),
    "| Total Expected Loss: $", format(sum(portfolio$expected_loss), big.mark = ","),
     sprintf("(%.3f%% of EAD)\n\n", 100 * sum(portfolio$expected_loss) / sum(portfolio$ead)))

cat("=== Expected loss by rating ===\n")
by_rating <- aggregate(cbind(ead, expected_loss) ~ rating, data = portfolio, FUN = sum)
by_rating <- by_rating[match(pd_table$rating, by_rating$rating), ]
by_rating$el_rate_bps <- round(10000 * by_rating$expected_loss / by_rating$ead, 1)
print(by_rating, row.names = FALSE)

cat("\n=== Expected loss by segment ===\n")
by_segment <- aggregate(cbind(ead, expected_loss) ~ segment, data = portfolio, FUN = sum)
by_segment$el_rate_bps <- round(10000 * by_segment$expected_loss / by_segment$ead, 1)
print(by_segment[order(-by_segment$el_rate_bps), ], row.names = FALSE)
 
```

### 5. Chart: expected loss by rating

```{r el-chart, fig.width=8, fig.height=5}
barplot(by_rating$el_rate_bps, names.arg = by_rating$rating,
        main = "Expected Loss Rate by Rating (bps of EAD)",
        ylab = "EL (basis points)", col = "steelblue")
```

```{r export, include=FALSE}
png("el_by_rating.png", width = 800, height = 500)
barplot(by_rating$el_rate_bps, names.arg = by_rating$rating,
        main = "Expected Loss Rate by Rating (bps of EAD)",
        ylab = "EL (basis points)", col = "steelblue")
dev.off()
write.csv(portfolio, "portfolio.csv", row.names = FALSE)
```

## Part II: Monte Carlo Capital

```{r}
if (!exists("portfolio")) {
  portfolio <- read.csv("portfolio.csv")
}
n_loans <- nrow(portfolio)
```

```{r}
# Pre-compute per-loan inputs as vectors (fast inside the simulation loop)
pd_vec      <- portfolio$pd
lgd_vec     <- portfolio$lgd
ead_vec     <- portfolio$ead
default_thresh <- qnorm(pd_vec)   # asset-value level below which loan defaults
```

```{r}
# 1. The simulation
set.seed(123)
simulate_losses <- function(n_sims = 10000, rho = 0.20) {
  # rho = asset correlation. Basel uses 0.12-0.24 for corporates; 0.20 is a
  # reasonable single value for a stylized corporate book.
 
  
  losses <- numeric(n_sims)
 
  for (s in 1:n_sims) {
    M <- rnorm(1)                        # the economy this year (one draw)
    e <- rnorm(n_loans)                  # each borrower's own luck
    A <- sqrt(rho) * M + sqrt(1 - rho) * e   # asset values
    defaulted <- A < default_thresh          # TRUE/FALSE per loan
    losses[s] <- sum(ead_vec[defaulted] * lgd_vec[defaulted])
  }
  losses
}
 
cat("Running 10,000 scenarios... (~10-20 seconds)\n")
losses <- simulate_losses(n_sims = 10000, rho = 0.20)
```

```{r}

# 2. From loss distribution to capital
el_sim   <- mean(losses)                  # expected loss (simulated)
var_999  <- quantile(losses, 0.999)       # 99.9% Value-at-Risk
ecap     <- var_999 - el_sim              # economic capital = UL beyond EL
var_99   <- quantile(losses, 0.99)
 
cat("\n=== Loss distribution summary (rho = 0.20) ===\n")
cat(sprintf("Expected loss (mean):        $%s\n", format(round(el_sim),  big.mark = ",")))
cat(sprintf("Median year:                 $%s\n", format(round(median(losses)), big.mark = ",")))
cat(sprintf("99%% VaR   (1-in-100 year):   $%s\n", format(round(var_99),  big.mark = ",")))
cat(sprintf("99.9%% VaR (1-in-1000 year):  $%s\n", format(round(var_999), big.mark = ",")))
cat(sprintf("Economic capital (VaR - EL): $%s\n", format(round(ecap),    big.mark = ",")))
cat(sprintf("Capital as %% of EAD:         %.2f%%\n", 100 * ecap / sum(ead_vec)))
cat(sprintf("Worst simulated year:        $%s\n", format(round(max(losses)), big.mark = ",")))
 
# Sanity check: simulated mean should be close to Day 1's analytic EL
cat(sprintf("\nSanity check - analytic EL from Day 1: $%s (simulated mean should be close)\n",
            format(sum(round(pd_vec * lgd_vec * ead_vec)), big.mark = ",")))
```

```{r}
# 3. What does correlation do?
#    Same portfolio, same PDs - only rho changes.
cat("\n=== Correlation sensitivity (the tail-fattening effect) ===\n")
cat("rho | Expected Loss | 99.9% VaR | Economic Capital\n")
for (r in c(0.05, 0.20, 0.40)) {
  set.seed(42)  # same randomness each run -> differences are PURELY from rho
  l <- simulate_losses(n_sims = 5000, rho = r)
  cat(sprintf("%.2f | $%12s | $%12s | $%12s\n",
              r,
              format(round(mean(l)), big.mark = ","),
              format(round(quantile(l, 0.999)), big.mark = ","),
              format(round(quantile(l, 0.999) - mean(l)), big.mark = ",")))
}
cat("\nNotice: EL barely moves, but VaR/capital explodes with rho.\n")
cat("Correlation doesn't change AVERAGE losses - it changes how losses CLUSTER.\n")
```

```{r loss-dist-chart, fig.width=9, fig.height=5.5}
# 4. Charts
hist(losses / 1e6, breaks = 60, col = "steelblue", border = "white",
     main = "Simulated 1-Year Portfolio Loss Distribution (rho = 0.20)",
     xlab = "Loss ($ millions)", ylab = "Frequency")
abline(v = el_sim / 1e6,  col = "darkgreen", lwd = 3)
abline(v = var_999 / 1e6, col = "red",       lwd = 3)
legend("topright",
       legend = c(sprintf("Expected Loss: $%.1fM", el_sim / 1e6),
                  sprintf("99.9%% VaR: $%.1fM", var_999 / 1e6)),
       col = c("darkgreen", "red"), lwd = 3, bty = "n")
```

```{r export-files, include=FALSE}
png("loss_distribution.png", width = 900, height = 550)
hist(losses / 1e6, breaks = 60, col = "steelblue", border = "white",
     main = "Simulated 1-Year Portfolio Loss Distribution (rho = 0.20)",
     xlab = "Loss ($ millions)", ylab = "Frequency")
abline(v = el_sim / 1e6,  col = "darkgreen", lwd = 3)
abline(v = var_999 / 1e6, col = "red",       lwd = 3)
legend("topright",
       legend = c(sprintf("Expected Loss: $%.1fM", el_sim / 1e6),
                  sprintf("99.9%% VaR: $%.1fM", var_999 / 1e6)),
       col = c("darkgreen", "red"), lwd = 3, bty = "n")
dev.off()
write.csv(data.frame(sim = 1:length(losses), loss = losses),
          "simulated_losses.csv", row.names = FALSE)
```

## Part III: Stress Testing & Return on Capital

```{r}
if (!exists("portfolio")) portfolio <- read.csv("portfolio.csv")
n_loans <- nrow(portfolio)

pd_table <- data.frame(
  rating = c("AAA", "AA", "A", "BBB", "BB", "B", "CCC"),
  pd     = c(0.0001, 0.0002, 0.0006, 0.0018, 0.0075, 0.0380, 0.2500)
)
```

### 1. Severely adverse scenario

Stylized severely adverse scenario:

a) Ratings migrate DOWN 1 notch (recession -> credit quality deteriorates)
b) LGD rises +15 points (collateral values fall when everyone sells)
c) Correlation rises 0.20 -> 0.35 (in crises, everything moves together)

```{r}
apply_stress <- function(pf) {
  ratings_order <- pd_table$rating                # AAA ... CCC
  idx <- match(pf$rating, ratings_order)
  idx_stressed <- pmin(idx + 1, length(ratings_order))   # down 1 notch, floor at CCC

  pf$rating_stressed <- ratings_order[idx_stressed]
  pf$pd_stressed  <- pd_table$pd[idx_stressed]
  pf$lgd_stressed <- pmin(pf$lgd + 0.15, 1.0)            # +15pts, capped at 100%
  pf
}

portfolio <- apply_stress(portfolio)
```

### 2. Simulation engine (taking pd/lgd vectors as inputs)

```{r}
simulate_losses <- function(pd_vec, lgd_vec, ead_vec,
                            n_sims = 10000, rho = 0.20, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  thresh <- qnorm(pd_vec)
  n <- length(pd_vec)
  losses <- numeric(n_sims)
  for (s in 1:n_sims) {
    M <- rnorm(1)
    e <- rnorm(n)
    A <- sqrt(rho) * M + sqrt(1 - rho) * e
    d <- A < thresh
    losses[s] <- sum(ead_vec[d] * lgd_vec[d])
  }
  losses
}
```

Running both worlds with dedicated seeds:

```{r}
cat("Running baseline and stressed simulations...\n")
losses_base   <- simulate_losses(portfolio$pd, portfolio$lgd, portfolio$ead,
                                 rho = 0.20, seed = 123)
losses_stress <- simulate_losses(portfolio$pd_stressed, portfolio$lgd_stressed,
                                 portfolio$ead, rho = 0.35, seed = 123)

summarize <- function(l, label) {
  data.frame(
    scenario = label,
    EL_M     = round(mean(l) / 1e6, 1),
    VaR999_M = round(quantile(l, 0.999) / 1e6, 1),
    ECap_M   = round((quantile(l, 0.999) - mean(l)) / 1e6, 1),
    ECap_pct_EAD = round(100 * (quantile(l, 0.999) - mean(l)) / sum(portfolio$ead), 2)
  )
}

cat("\n=== Baseline vs Severely Adverse ===\n")
comparison <- rbind(summarize(losses_base, "Baseline"),
                    summarize(losses_stress, "Severely adverse"))
print(comparison, row.names = FALSE)

cat("\nCapital adequacy read: if the bank holds capital sized to the BASELINE,\n")
cat("the stressed requirement shows the shortfall a stress test would reveal.\n")
```

The table above compares economic capital requirements under two parameterizations:
the same 99.9% standard applied to a baseline world and a stressed world.
Capital demand rises from $76.7M (5.4% of EAD) to $287.5M (20.4%),
meaning a bank capitalized for baseline conditions could not meet the
same solvency standard if stressed conditions became the new normal.

How this differs from regulatory DFAST: my comparison shows how the capital
requirement changes under stress. DFAST asks a different question: whether the
bank's current capital can absorb stressed losses (~$43M/year here) over a
nine-quarter horizon while staying above minimum ratios, a test this portfolio
would pass. In short: mine informs capital planning; DFAST is the regulatory
pass/fail.

Limitations: the one-notch downgrade and single-year horizon are simplifications;
real DFAST transmits Fed macro variables through satellite models over nine
quarters.

### 3. Return on Capital by rating bucket

Income = spread earned on the loan (stylized: wider spread for riskier).
ROC = (income - expected loss) / capital allocated to that bucket.
Stylized annual credit spreads by rating (bps), roughly market-shaped.

```{r}
spread_bps <- c(AAA = 50, AA = 65, A = 90, BBB = 140, BB = 275, B = 450, CCC = 900)

portfolio$spread   <- spread_bps[portfolio$rating] / 10000
portfolio$income   <- portfolio$ead * portfolio$spread
portfolio$el       <- portfolio$pd * portfolio$lgd * portfolio$ead

# Capital per loan: allocating total baseline economic capital in proportion to
# each loan's standalone unexpected-loss proxy: EAD * LGD * sqrt(PD*(1-PD))
ul_proxy <- portfolio$ead * portfolio$lgd * sqrt(portfolio$pd * (1 - portfolio$pd))
ecap_total <- quantile(losses_base, 0.999) - mean(losses_base)
portfolio$capital <- ecap_total * ul_proxy / sum(ul_proxy)

roc <- aggregate(cbind(income, el, capital, ead) ~ rating, data = portfolio, FUN = sum)
roc <- roc[match(pd_table$rating, roc$rating), ]
roc$roc_pct <- round(100 * (roc$income - roc$el) / roc$capital, 1)
roc[, c("income","el","capital","ead")] <- round(roc[, c("income","el","capital","ead")] / 1e6, 2)

cat("\n=== Return on Capital by rating ($M) ===\n")
print(roc[, c("rating","ead","income","el","capital","roc_pct")], row.names = FALSE)

cat("high ROC = earns its capital. Watch which buckets are best -\n")
cat("it's usually NOT the safest ones (thin spreads) NOR the riskiest (EL eats\n")
cat("the spread, capital charge huge). The middle of the book pays the bills.\n")
```

### 4. Chart: the two loss distributions overlaid

```{r stress-overlay, fig.width=9, fig.height=5.5}
plot_stress_overlay <- function() {
  brks <- seq(0, max(losses_stress) / 1e6 + 10, by = 3)
  h_base   <- hist(losses_base / 1e6,   breaks = brks, plot = FALSE)
  h_stress <- hist(losses_stress / 1e6, breaks = brks, plot = FALSE)
  ylim <- c(0, max(h_base$counts, h_stress$counts))

  plot(h_base, col = rgb(0.27, 0.51, 0.71, 0.6), border = "white", ylim = ylim,
       main = "Loss Distribution: Baseline vs Severely Adverse",
       xlab = "Loss ($ millions)", ylab = "Frequency")
  plot(h_stress, col = rgb(0.85, 0.33, 0.31, 0.55), border = "white", add = TRUE)
  abline(v = quantile(losses_base, 0.999)/1e6,   col = "steelblue4", lwd = 3, lty = 2)
  abline(v = quantile(losses_stress, 0.999)/1e6, col = "red3",       lwd = 3, lty = 2)
  legend("topright",
         legend = c("Baseline", "Severely adverse",
                    "99.9% VaR (baseline)", "99.9% VaR (stressed)"),
         fill   = c(rgb(0.27,0.51,0.71,0.6), rgb(0.85,0.33,0.31,0.55), NA, NA),
         border = NA,
         col    = c(NA, NA, "steelblue4", "red3"),
         lty    = c(NA, NA, 2, 2), lwd = c(NA, NA, 3, 3), bty = "n")
}

plot_stress_overlay()   # shows inline in a notebook chunk

png("stress_overlay.png", width = 950, height = 550)
plot_stress_overlay()
dev.off()
write.csv(comparison, "stress_comparison.csv", row.names = FALSE)
cat("\nSaved: stress_overlay.png, stress_comparison.csv\n")
```
